MINST Kaggle Digit Recognizer: Contrasting the Random Forest and MLP Neural Network

### 2019-04-21 paper/paper.tex
Paper changes
Add files via upload

\section{Random Forest}
A Random Forest (RF) is a common starting algorithm for those new to machine learning. This is in part due to it's easy of use, but also due to it's flexibility. While in this assignment it is used for classification, it can be also be used in regression problems [2]. Each tree is formed from a bootstrap sample (random sampling with replacement) then grown very similarly to a decision tree. Our random forest is a ``highly random'' random forest which splits data with no discretion to promote model volatility.\verb+Sklearn+'s \verb+RandomForestClassifier+ combines the probabilistic prediction of each tree, as this has been shown to preform better than the traditional approach of picking a classification based on a majority vote [4,5].

These models are generally very fast to train, but depending on the complexity, can suffer from slow run-time performance. Generally speaking, overfitting can be negated by simple adding more trees to your model [2]. This approach eventually results in diminishing returns, and for high-dimensional problems such as digit recognition, is especially impractical [5].

@@

\subsection{Cross-Validation and Results}
Training and cross-validation was done using \verb+model_selection.GridSearchCV+. In order to visualize the results, Figure 2 contrasts the mean CVS and prediction rates of the training and test datasets. The $\log_2(minleafsize)$ which inversely measures the complexity suggests the more complex models preform outstandingly against the training data but have little impact on the testing data, a symptom of overfitting. Adding trees has a similar positive affect on both datasets, however, also decreases exponentially as more trees are added. These results suggest that our model suffers from extreme levels of overfitting. As expected, adding trees does improve our models predictive power but isn't the most practical for our dataset. Our Kaggle submission of 300 trees and a minimum leaf number of 2 gave a prediction rate of 96.6\%

@@

\subsection{Implementation}
In addition to contrasting GD algorithms, the number of hidden nodes were varied between 128, 256, and 512; the number of hidden layers were either 1 or 2; the regularization term $\alpha$ was varied at 0.5, 0.1, 0.001, and 0.0001. The other regularization parameter's for Adam GD were kept at the developer's (Kingma and Lei Ba) recommended defaults: $\epsilon = 10^{-8}$, $\beta_1=0.9$, and $\beta_2=0.999$ [3]. While the learning rate was not varied, an adaptive rate was chosen which divides the current learning rate by 5 with a starting value of $0.001$ [4]. One the code was prepared, training took approximately 46 minutes per model. Total training time took approximately 20 hours.

\subsection{Cross-Validation and Results}
Training and cross-validation was done using \verb+model_selection.GridSearchCV+. In order to visualize the results, Figure 2 contrasts the mean CVS and prediction rates of the training and test datasets. A measure of complexity is given by $-log_{10}$ meaning lower $\alpha$ values correspond to higher complexity. As expected, Adam gradient descent was the clear winner across the board. Note that the smaller $\alpa$ value (that is the larger $-\log_{10}(\alpha)$ value) almost consistently improves the prediction rate in (A), but for (B) in some cases causes severe prediction penalties. Examples most obvious example of this is at 1 HL and 256 nodes in (B). Our "winning" model with 1 HL and 512 nodes preformed rather consistently between training and testing datasets. The Kaggle submission of this model gave a prediction rate of 97.90\%.

@@

\section{Conclusions}
As expected, our MLP was better suited for handwritten digit recognition. It is possible that data processing to reduce dimensionality would help the RF preform better, however computer vision problems are still better left in the hands of NN's and similar models. One advantage of the RF is it's training time was approximately $1/60^{th}$ of our MLP's, however adding addition trees would begin to bridge this gap and further improve it's classification accuracy.

>>>>

\section{Random Forest}
A Random Forest (RF) is a common starting algorithm for those new to machine learning. This is in part due to it's ease of use, but also due to it's flexibility. While in this assignment it is used for classification, it can be also be used in regression problems [2]. Each tree is formed from a bootstrap sample (random sampling with replacement) then grown very similarly to a decision tree. Our random forest is a ``highly random'' random forest which splits data with no discretion to promote model volatility.\verb+Sklearn+'s \verb+RandomForestClassifier+ combines the probabilistic prediction of each tree, as this has been shown to preform better than the traditional approach of picking a classification based on a majority vote [4,5].

These models are generally very fast to train, but depending on the complexity, can suffer from slow run-time performance. Generally speaking, overfitting can be negated by simply adding more trees to your model [2]. This approach eventually results in diminishing returns, and for high-dimensional problems such as digit recognition, is especially impractical [5].

@@

\subsection{Cross-Validation and Results}
Training and cross-validation was done using \verb+model_selection.GridSearchCV+. In order to visualize the results, Figure 2 contrasts the mean CVS and prediction rates of the training and test datasets. The $\log_2(\text{minleafsize})$ which inversely measures the complexity suggests the more complex models preform outstandingly against the training data but have little impact on the testing data, a symptom of overfitting. Adding trees has a similar positive affect on both datasets, however, also decreases exponentially as more trees are added. These results suggest that our model suffers from extreme levels of overfitting. As expected, adding trees does improve our models predictive power but isn't the most practical for our dataset. Our Kaggle submission of 300 trees and a minimum leaf number of 2 gave a prediction rate of 96.6\%

@@

\subsection{Implementation}
In addition to contrasting GD algorithms, the number of hidden nodes were varied between 128, 256, and 512; the number of hidden layers were either 1 or 2; the regularization term $\alpha$ was varied at 0.5, 0.1, 0.001, and 0.0001. The other regularization parameters for Adam GD were kept at the developer's (Kingma and Lei Ba) recommended defaults: $\epsilon = 10^{-8}$, $\beta_1=0.9$, and $\beta_2=0.999$ [3]. While the learning rate was not varied, an adaptive rate was chosen which divides the current learning rate by 5 with a starting value of $0.001$ [4]. Once the code was prepared, training took approximately 46 minutes per model. Total training time took approximately 20 hours.

\subsection{Cross-Validation and Results}
Training and cross-validation was done using \verb+model_selection.GridSearchCV+. In order to visualize the results, Figure 2 contrasts the mean CVS and prediction rates of the training and test datasets. A measure of complexity is given by $-\log_{10}(\alpha)$ meaning lower $\alpha$ values correspond to higher complexity. As expected, Adam gradient descent was the clear winner across the board. Note that the smaller $\alpha$ values (that is the larger $-\log_{10}(\alpha)$ values) almost consistently improves the prediction rate in (A), but for (B) in some cases causes severe prediction penalties. The most obvious example of this is at 1 HL and 256 nodes in (B). Our "winning" model with 1 HL and 512 nodes preformed rather consistently between training and testing datasets. The Kaggle submission of this model gave a prediction rate of 97.90\%.

@@

\section{Conclusions}
As expected, our MLP was better suited for handwritten digit recognition. It is possible that data processing to reduce dimensionality would help the RF preform better, however computer vision problems are still better left in the hands of NN's and similar models. One advantage of the RF is it's training time was approximately $1/60^{th}$ of our MLP's, however adding additional trees would begin to bridge this gap and further improve it's classification accuracy.
